---
title: "Assignment1"
author: "Xingguang Li"
date: "December 25, 2017"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, warning = FALSE, message = FALSE)
```

## Introduction 
In this assignment, the activity data set is explorerd and analyzed. 

## Data loading and preprocessing
The csv file is loaded along with the formatting of time variables.
```{r load-data}
activity <- read.csv("activity.csv",head=TRUE)
activity$date <- as.Date(activity$date,format="%Y-%m-%d")
head(activity,5)
str(activity)
```

## Mean total number of steps taken per day
### Aggregate the total number of steps per day
To obtain the daily total steps, the step data is aggregated for each day.

The new data set is stored in activity-daily.

```{r}
library(dplyr)
activity_daily <- aggregate(steps~date,activity,sum,na.rm=TRUE)
```
### Histogram of total number of steps per day
Now the histogram of total number of steps per day is plotted.
```{r Hist, fig.height=3, fig.width=6,fig.align='center'}
hist(activity_daily$steps,breaks = 50, main = "Total number of steps per day", xlab = "Steps")
```

### Mean and median of total number of steps per day

The mean and median of the total number of steps per day are calculated as below:
```{r}
mean(activity_daily$steps)
median(activity_daily$steps)
```

## average daily activity pattern
Now we investigate the daily activity pattern.

To do this we first aggregate the number of steps for each 5 min over all days.

```{r}
activity_5min <- aggregate(steps~interval,activity,sum,na.rm=TRUE)
```

### Time series plot
Here is a time series plot presenting the number of steps for each 5 min over all days.
```{r TS1, fig.height=3, fig.width=6,fig.align='center'}
plot(activity_5min$interval,activity_5min$steps,type="l",main="Number of steps for each 5 min", xlab="Time", ylab="Steps")
```

### The interval with the maximum number of steps 
The interval that has the maximal number of steps on average is calculated as below:
```{r}
activity_5min %>% subset(steps == max(steps))
```
So the interval 835 has the maximal average number of steps.

## Imputing missing values
### Number of missing values
```{r}
nrow(activity %>% filter(is.na(steps))) 
```
### Median Imputation
The imputation technique adopted here is the median imputation for each interval.
```{r,cache=TRUE}
steps_median <- activity %>% aggregate(steps~interval,data=.,median,na.rm=TRUE)
activity$steps_impute <- ifelse(is.na(activity$steps),steps_median[which(activity$interval==steps_median$interval),"steps"],activity$steps)
```

### Histogram of total number of steps per day
```{r Hist2,fig.height=3, fig.width=6,fig.align='center'}
activity_daily_impute <- aggregate(steps_impute~date,activity,sum,na.rm=TRUE)
hist(activity_daily_impute$steps_impute,breaks=50,main = "Total number of steps per day", xlab = "Steps")
```

### Mean and median of total number of steps per day
The mean and median of the total number of steps per day after imputation are calculated as below:
```{r}
mean(activity_daily_impute$steps_impute)
median(activity_daily_impute$steps_impute)
```
In comparison with the values before imputation, the imputed mean and median go down because of the imputed values regressing to the median. 

### Difference between weekdays and weekends
Determine weekdays and weekends of the imputed data set
```{r}
activity <- activity %>% mutate(date_wk=weekdays(date))
activity$date_wk <- ifelse(activity$date_wk %in% c("Sunday","Saturday"),"Weekend","Weekday")
activity$date_wk <- as.factor(activity$date_wk)
```
Plot the corresponding time series using lattice plot.
```{r TS2,fig.height=4, fig.width=6,fig.align='center'}
library(lattice)
xyplot(steps_impute ~ interval|date_wk, data=activity, type='l',layout=c(1,2),ylab="Steps", xlab="Time", main="Number of steps over time")
```

It can be observed that people start to walk earlier on weekdays than weekends.